Quantitative, Multispecies Monitoring at a Continental Scale

### 2025-02-14 Manuscript_plain_AOS.tex
update main text

\begin{center}
\begin{tabular}{ll}
1 & School of Marine and Environmental Affairs, University of Washington, Seattle, Washington, USA \\
2 & Northwest Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA \\
3 & Museo Nacional de Ciencias Naturales, Madrid, Spain \\
\hline
\textbf{*} & \textbf{Gledis Guri}, lead and corresponding author \textbf{gguri@uw.edu}
\end{tabular}
\end{center}

@@

\section*{Author Contributions}
AOS, KMN, RPK, MS, and KP designed the research;\\
MRS, AR-L and AW did the laboratory work; \\
PB, RPK, and GG did the bioinformatics;\\
GG, OL, RPK, and eAOS analyzed the data;\\
GG, OL, RPK, and AOS interpreted the results;\\
All authors provided edits to the manuscript

[Please check that this is correct and add/edit as needed XXX]

@@

\section{Introduction}
Environmental DNA (eDNA, residual genetic material sampled from water, soil, or air) is an increasingly common tool for non-invasively sampling marine and aquatic ecological communities, valuable for estimating species occurrence \cite{veilleux2021}, biodiversity (i.e., Richness, \cite{muenzel2024}), and genetic structure \cite{andres2023a}. However, eDNA has not been widely used for quantitative abundance estimation, upon which many practical management and conservation applications rely (but see \cite{shelton2022,guri2024a,stoeckle2022a}). The use of eDNA for quantification is supported both conceptually -- where there is more of a species, more cells are shed in the environment, and thus there is more DNA -- and empirically (e.g. there are strong linkages between single-species eDNA measures of in aquaria \cite{jo2019b,ledger2024}, rivers and streams \cite{pont2024}, estuaries and nearshore habitats \cite{dibattista2022, baetscher2024}, and coastal oceans \cite{maes2023, shelton2019}). However, uses of quantitative eDNA observations in management applications have significantly lagged behind their use in occurrence and biodiversity applications.

@@

\section{Results}
We find spatial patterns of DNA concentrations reflect the distributions of commercially important coastal-pelagic fish species in three dimensions and across ten degrees of latitude in a single year of sampling in 2019 (Fig. \ref{fig:fig1}: Northern Anchovy, Pacific Sardine; \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}, Pacific Herring and Jack Mackerel). For other high-biomass species otherwise lacking well-described distributions, we provide the first species distribution models (Fig. \ref{fig:fig1}, \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}). We find that the estimated eDNA concentrations are correlated among species with similar traits and habitat preferences (Fig. \ref{fig:fig2}). Emergent patterns of species richness and estimated eDNA aggregation correspond with known ecological drivers, evidence that eDNA traces can reliably capture ecologically meaningful patterns (Fig. \ref{fig:fig3}). Particularly for forage fish species – which are critical high-biomass links in marine food webs, but difficult to survey with traditional methods \cite{Pikitch2012} – molecular methods offer a tractable and quantitative index of abundance over large spatial scales.

@@

The eDNA patterns for northern lampfish (Stenobrachius leucopsarus) -- an abundant myctophid that migrates from 400-700m up to 20-200m at night \cite{moku2000,suntsov2008} -- are again consistent with the ecology of the species, having notably lower concentrations in surface samples. Owing to the diel vertical migration of the northern lampfish, the signal of eDNA for this species is observed across depths.

@@

\begin{figure}[tbhp] 
\centering
\includegraphics[width=8.5cm]{plots/2_Figure_2.jpg}  
\caption{Correlation plot describing species co-abundance and co-occurance. Spearman’s rho correlation coefficient are expressed in numbers and colors ranging from -1 (red) negatively correlated to 1 (blue) positively correlated. Species are ordered in known ecological groups (surface, midwater, and deepwater) to visually indicate the gradual change of species co-occurance.}
\label{fig:fig2}
\end{figure}

@@

\section{Discussion}
We report species-specific DNA concentrations smoothed over ten degrees of latitude and across the continental shelf to 500m. We reveal distributional pattern for 12 species ranging from the iconic coastal pelagic species of the northeast Pacific to understudied-but-abundant deepwater species, and our results suggest clear pathways for the broader use of eDNA in both fisheries and ecosystem management.

@@

\section{Methods}
We summarize all methodological components and provide details for all sampling, laboratory, and statistical methods in the \href{SI_Appendix.pdf}{SI Appendix}.

@@

\begin{figure*}[tbhp] 
\centering
\includegraphics[width=16cm]{plots/DAG-Ole.pdf} 
\caption{Schematic for the statistical models to produce species distributions. Three sets of observations (orange) were estimated simultaneously to provide estimated eDNA concentrations twelve fish species (blue) from which we derive smoothed predictions of DNA concentration (green). Mock communities for metabarcoding (yellow) were estimated externally and provided estimated $\alpha_s$ that were used in the joint model. Notation for parameters, subscripts, and observations are indicated in gray. See \href{SI_Appendix.pdf}{SI Appendix} for full model breakdown.}
\label{fig:DAG2}
\end{figure*}

@@

We follow \cite{shelton2023} and construct a compositional model for the metabarcoding data from 371 of the water samples analyzed for Pacific hake qPCR. We use the metabarcoding observations of the 8 mock communities to calibrate the relative amplification efficiency of the 12 focal species (plus hake, a 13$^{th}$ species). In the absence of qPCR data, this model would produce estimates of proportional DNA contributions for each water sample. But for each sample in which we have non-zero estimates of hake DNA concentration from qPCR, and non-zero estimate of hake proportional contribution from metabarcoding (\href{SI_Appendix.pdf}{SI Appendix Fig. S4}), we can expand from the DNA concentration of hake to provide an estimate of the DNA concentration of the 12 remaining focal species \cite{allan2023}. Our model generates estimates of DNA concentration (copies/L) for each species. Thus hake is a lynchpin that allows us to provide estimates of DNA concentration for all other species; we refer to hake as our “reference species”. Hake was present in 86.7\% of samples via qPCR, and 67.7\% of samples via metabarcoding, making it an appropriate and important reference species. To enhance sample coverage -- by increasing the number of non-zero estimates of link species -- and improve model accuracy, multiple single-species qPCR assays can be integrated as additional sources of DNA concentration. However, this approach must be applied thoughtfully, ensuring that selected qPCR targeted species have minimal co-occurrence, like Pacific sardine or northern anchovy in our case (\href{SI_Appendix.pdf}{SI Appendix Fig. S5}). Nevertheless, samples lacking hake detections here were sporadically distributed across the study area (\href{SI_Appendix.pdf}{SI Appendix Fig. S4}), and deserve further attention in future work.

@@

The above statistical model (Fig. \ref{fig:DAG2}) provides predictions of species-specific DNA concentrations (D) at each location and depth sampled ($d$ = 0, 50, 150, 300, or 500 m). To generate smooth maps for each species, we used the sdmTMB package \cite{anderson2022} to generate smooth maps of DNA concentration for each species independently (Fig. \ref{fig:DAG2}, green box).

>>>>

\begin{center}
\begin{tabular}{ll}
1 & School of Marine and Environmental Affairs, University of Washington, Seattle, Washington, USA \\
2 & Conservation Biology Division, Northwest Fisheries Science Center,  NOAA Fisheries, \\
 & \qquad 2715 Montlake Blvd E. Seattle, Washington, USA \\
3 & Museo Nacional de Ciencias Naturales, Madrid, Spain \\
\hline
\textbf{*} & \textbf{Gledis Guri}, lead and corresponding author \textbf{gguri@uw.edu}
\end{tabular}
\end{center}

@@

\section*{Author Contributions}
AOS, KMN, RPK, MS, and KP designed the research;\\
MRS, AR-L and AW did the laboratory work; \\
PB, RPK, and GG did the bioinformatics;\\
GG, OL, RPK, and eAOS analyzed the data;\\
GG, OL, RPK, and AOS interpreted the results;\\
All authors provided edits to the manuscript \\
Please check that this is correct and add/edit as needed XXX

@@

\section*{Introduction}
Environmental DNA (eDNA, residual genetic material sampled from water, soil, or air) is an increasingly common tool for non-invasively sampling marine and aquatic ecological communities, valuable for estimating species occurrence \cite{veilleux2021}, biodiversity (i.e., Richness, \cite{muenzel2024}), and genetic structure \cite{andres2023a}. However, eDNA has not been widely used for quantitative abundance estimation, upon which many practical management and conservation applications rely (but see \cite{shelton2022,guri2024a,stoeckle2022a}). The use of eDNA for quantification is supported both conceptually -- where there is more of a species, more cells are shed in the environment, and thus there is more DNA -- and empirically (e.g. there are strong linkages between single-species eDNA measures of in aquaria \cite{jo2019b,ledger2024}, rivers and streams \cite{pont2024}, estuaries and nearshore habitats \cite{dibattista2022, baetscher2024}, and coastal oceans \cite{maes2023, shelton2019}). However, uses of quantitative eDNA observations in management applications have significantly lagged behind their use in occurrence and biodiversity applications.

@@

\section*{Results}
We find spatial patterns of DNA concentrations reflect the distributions of commercially important coastal-pelagic fish species in three dimensions and across ten degrees of latitude in a single year of sampling in 2019 (Fig. \ref{fig:fig1}: Northern Anchovy, Pacific Sardine; \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}, Pacific Herring and Jack Mackerel). For other high-biomass species otherwise lacking well-described distributions, we provide the first species distribution models (Fig. \ref{fig:fig1}, \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}). We find that the estimated eDNA concentrations are correlated among species with similar traits and habitat preferences (Fig. \ref{fig:fig2}). Emergent patterns of species richness and estimated eDNA aggregation correspond with known ecological drivers, evidence that eDNA traces can reliably capture ecologically meaningful patterns (Fig. \ref{fig:fig3}). Particularly for forage fish species – which are critical high-biomass links in marine food webs, but difficult to survey with traditional methods \cite{Pikitch2012} – molecular methods offer a tractable and quantitative index of abundance over large spatial scales.

@@

The eDNA patterns for northern lampfish (\textit{Stenobrachius leucopsarus}) -- an abundant myctophid that migrates from 400-700m up to 20-200m at night \cite{moku2000,suntsov2008} -- are again consistent with the ecology of the species, having notably lower concentrations in surface samples. Owing to the diel vertical migration of the northern lampfish, the signal of eDNA for this species is observed across depths.

@@

\begin{figure}[tbhp] 
\centering
\includegraphics[width=10cm]{plots/2_Figure_2.jpg}  
\caption{Correlation plot describing species co-abundance and co-occurance. Spearman’s rho correlation coefficient are expressed in numbers and colors ranging from -1 (red) negatively correlated to 1 (blue) positively correlated. Species are ordered in known ecological groups (surface, midwater, and deepwater) to visually indicate the gradual change of species co-occurance.}
\label{fig:fig2}
\end{figure}

@@

\section*{Discussion}
We report species-specific DNA concentrations smoothed over ten degrees of latitude and across the continental shelf to 500m. We reveal distributional pattern for 12 species ranging from the iconic coastal pelagic species of the northeast Pacific to understudied-but-abundant deepwater species, and our results suggest clear pathways for the broader use of eDNA in both fisheries and ecosystem management.

@@

\section*{Methods}
We summarize all methodological components and provide details for all sampling, laboratory, and statistical methods in the \href{SI_Appendix.pdf}{SI Appendix}.

@@

\begin{figure*}[tbhp] 
\centering
\includegraphics[width=16cm]{plots/DAG-Ole.pdf} 
\caption{Schematic for the statistical models to produce species distributions. Three sets of observations (orange) were used simultaneously to provide estimates of eDNA concentration for  twelve fish species (blue) from which we derive smoothed predictions of DNA concentration (green). Mock communities for metabarcoding (yellow) were estimated externally and provided estimated $\alpha_s$ that were used in the joint model. Notation for parameters, subscripts, and observations are indicated in gray. See \href{SI_Appendix.pdf}{SI Appendix} for full model description.}
\label{fig:DAG2}
\end{figure*}

@@

We follow \cite{shelton2023} and construct a compositional model for the metabarcoding data from 371 of the water samples analyzed for Pacific hake qPCR. We use the metabarcoding observations of the 8 mock communities to calibrate the relative amplification efficiency of the 12 focal species (plus hake, a 13$^{th}$ species). In the absence of qPCR data, this model would produce estimates of proportional DNA contributions for each water sample. But for each sample in which we have non-zero estimates of hake DNA concentration from qPCR, and non-zero estimate of hake proportional contribution from metabarcoding (\href{SI_Appendix.pdf}{SI Appendix Fig. S4}), we can expand from the DNA concentration of hake to provide an estimate of the DNA concentration of the 12 remaining focal species \cite{allan2023}. Our model generates estimates of DNA concentration (copies/L) for each species. Thus hake is a lynchpin that allows us to provide estimates of DNA concentration for all other species. Hake was present in 86.7\% of samples via qPCR, and 67.7\% of samples via metabarcoding, making it an appropriate and important reference species. To enhance sample coverage -- by increasing the number of non-zero estimates of link species -- and improve model accuracy, multiple single-species qPCR assays can be integrated as additional sources of DNA concentration. However, this approach must be applied thoughtfully, ensuring that selected qPCR targeted species have minimal co-occurrence, like Pacific sardine or northern anchovy in our case (\href{SI_Appendix.pdf}{SI Appendix Fig. S5}). Nevertheless, samples lacking hake detections here were sporadically distributed across the study area (\href{SI_Appendix.pdf}{SI Appendix Fig. S4}), and deserve further attention in future work.

@@

The above statistical model (Fig. \ref{fig:DAG2}) provides predictions of species-specific DNA concentrations (\textit{D}) at each location and depth sampled ($d$ = 0, 50, 150, 300, or 500 m). To generate smooth maps for each species, we used the sdmTMB package \cite{anderson2022} to generate smooth maps of DNA concentration for each species independently (Fig. \ref{fig:DAG2}, green box).
